Multi-stage programming with functors and monads: eliminating abstraction overhead from generic code

### 2007-06-29 SCP2008/paper/metamonads.tex
Updated for the new code. Added notes

\begin{figure}
\begin{code}
type ('v,'s,'w) monad = 's -> ('s -> 'v -> 'w) -> 'w
let ret (a :'v) : ('v,'s,'w) monad = fun s k -> k s a
let bind a f = fun s k -> a s (fun s' b -> f b s' k)
let fetch s k = k s s  and  store v s k = k (v::s) ()

let k0 s v = v
let runM m = m [] k0

let l1 f = fun x -> perform \{ t <-- x; f t\}
let l2 f = fun x y -> perform \{ tx <-- x; ty <-- y; f tx ty\}

@@

let ifL test th el = ret .< if .~test then .~th else .~el >.
let ifM test th el = fun s k ->
  k s .< if .~(test s k0) then .~(th s k0) else .~(el s k0) >.
\end{code}
\caption{Our monad}\label{ourmonad}
\end{figure}

@@

In Figure~\ref{ourmonad}, the monad (yielding values of the type |v|)
is implemented as a function of two
arguments: the state (of type |s|) and the continuation. The
continuation receives the current state and the value, and
yields the answer of the type |w|.  The monad is polymorphic over the
three type parameters.  Other implementations are
possible. Except for the code in Figure~\ref{ourmonad}, the rest of our code
treats the monad as a truly abstract data type. The implementation of
the basic
monadic operations |ret| and |bind| is conventional and clearly
satisfies the monadic laws. Other monadic operations
construct computations that do have specific effects.  Operations |fetch| and
|store v| construct computations that read and write the state. In our
case the state is a list (of polymorphic variants), which models an
open discriminated union, as we shall see later.

@@

Finally, |runM| runs our monad, that is, performs the computation of
the monad and returns its result, which in our case is the code
expression. We run the monad by passing it the initial state and the
initial continuation |k0|. We can now re-write our |param_code2|
example of the previous section as |param_code3|.
\begin{code}
let param_code3 plus one =
  let gen x y = bind (retN (plus y one)) (fun ce -> 
                ret (plus ce (plus x ce)))
  in .<fun x y -> .~(runM (gen .<x>. .<y>.))>.
\end{code}
% param_code3 plus one;;
%
That does not seem like much of an improvement. With the help of
camlp4 pre-processor, we introduce the |perform|-notation \cite{metamonadsURL},
patterned after the |do|-notation of Haskell. 
\begin{code}
let param_code4 plus one =
  let gen x y = mdo \{ ce <-- retN (plus y one);
                      ret (plus ce (plus x ce)) \}
  in .<fun x y -> .~(runM (gen .<x>. .<y>.))>.
\end{code}
The function
|param_code4|, written in the |perform|-notation, is equivalent to
|param_code3| -- in fact, the camlp4 preprocessor will convert the
former into the latter. And yet, |param_code4| looks far more
conventional, as if it were indeed in direct style.

\subsection{Generating control statements}
We can write operations that generate code other than let-statements,
e.g., conditionals: see |ifL| in Figure~\ref{ourmonad}. The function |ifL|, 
albeit straightforward, is not as general as we wish: its arguments are
already generated pieces of code rather than monadic values. We
``lift it'', see |ifM'|. 
\begin{code}
let ifM' test th el = perform \{
  testc <-- test; thc <-- th; elc <-- el;
  ifL testc thc elc\}
let gen a i = ifM' (ret .<(.~i) >= 0>.) 
                   (retN .<Some (.~a).(.~i)>.) (ret .<None>.)
 in .<fun a i -> .~(runM (gen .<a>. .<i>.))>.
\evalresult{.<fun a_1 i_2 -> \protect\newline
let t_3 = (Some a_1.(i_2)) in if (i_2 >= 0) then t_3 else None>.}
let gen a i = ifM (ret .<(.~i) >= 0>.) 
                  (retN .<Some (.~a).(.~i)>.) (ret .<None>.)
 in .<fun a i -> .~(runM (gen .<a>. .<i>.))>.
\evalresult{.<fun a_1 i_2 -> 
if (i_2 >= 0) then let t_3 = (Some a_1.(i_2)) in t_3 else None>.}
\end{code}
%
We define functions |l1|,
|l2|, |l3| (analogues of |liftM|, |liftM2|, |liftM3| of Haskell) 
to make such a lifting generic. However we also need
another |ifM| function, with the same
interface (see Figure~\ref{ourmonad}).
The difference between them is
apparent: in the code above with |ifM'|, the let-insertion
happened \emph{before} the if-expression, that is, before the test that
the index |i| is positive. If |i| turned out
negative, |a.(i)| would generate an out-of-bound array access
error. On the other hand, the code with |ifM| accesses the array only
when we have verified that the index is non-negative. This example
makes it clear that the code generation (such as the one in |retN|) is 
truly an effect and we have to be clear about the sequencing of
effects when generating control constructions such as conditionals.
The form |ifM| handles such effects correctly. We
need similar operators for other OCaml control forms: for generating
case-matching statements and |for|- and |while|-loops.
\oleg{put them here?}

@@

The simplest parametrization is to make the domain abstract. As it
turns out, we need the following to exist in our domains: $0$, $1$,
$+$, $*$, (unary and binary) $-$, at least \emph{exact} division,
normalization, and potentially a relative size measure. The simplest
case of such domain abstraction is |param_code1|.
There, code-generators such as |plus| and |one|
were passed as arguments. We need far more than
two parameters, so we have to group them. Instead of the grouping
offered by regular records, we use OCaml \emph{structures} (i.e.,
modules)
so we can take advantage of extensibility, type abstraction and constraints,
and especially parameterized structures (\emph{functors}).
We define the type of the domain, the signature |DOMAIN|, which
different domains must satisfy:
\begin{figure}[h]
\begin{code}
module type DOMAIN = sig
  type v    type 'a vc = ('a,v) code
  type kind (* Field or Ring ? *)
  val zero : 'a vc   val one : 'a vc
  val plus : 'a vc -> 'a vc -> ('a vc, 's, 'w) monad
  (* times, minus, uminus, div elided for brevity *)
  val better_than : ('a vc -> 'a vc -> 
      (('a,bool) code, 's, 'w) monad) option
  val normalizerf : (('a,v -> v) code ) option
end 
module IntegerDomain : DOMAIN = struct
  type v = int  type kind = domain_is_ring
  type 'a vc = ('a,v) code
  let zero = .< 0 >.  and one = .< 1 >. 
  let plus x y = ret .<.~x + .~y>. 
  let better_than = Some (fun x y -> ret .<abs .~x > abs .~y >. )
  let normalizerf = None
  ...
end
\end{code}
\end{figure}

@@

\begin{code2}
module type OUTPUT = sig
  type contr  type res
  module D : DETERMINANT   module R : RANK   module P : TRACKPIVOT
  val make_result : ('a,contr) code -> 
    (('a,res) code,
     [> `TDet of 'a D.lstate | `TRan of 'a R.lstate | `TPivot of 'a P.lstate]
       list, ('a,'w) code) monad
end
module OutDetRank(Dom:DOMAIN)(C: CONTAINER2D)
    (Det : DETERMINANT with type indet = Dom.v and type outdet = Dom.v)
    (Rank : RANK) = struct
  module Ctr = C(Dom)
  type contr = Ctr.contr
  type res = contr * Det.outdet * int
  module D = Det   module R = Rank   module P = DiscardPivot
  let make_result b = perform \{ det  <-- D.fin ();  rank <-- R.fin ();
    ret .< ( .~b, .~det, .~rank ) >. \}
end
\end{code2}

@@

\begin{figure}
\begin{code2}
module Gen(Dom: DOMAIN)(C: CONTAINER2D)(PivotF: PIVOT)
          (Update: UPDATE with type baseobj = Dom.v and type ctr = C(Dom).contr)
          (Out: OUTPUT with type contr = C(Dom).contr and type D.indet = Dom.v 
                        and type 'a D.lstate = 'a Update.D.lstate) = struct
    module Ctr = C(Dom)
    module Pivot = PivotF(Dom)(C)(Out.D)
    let gen =
      let zerobelow b r c m n brc =
        let innerbody i = perform \{
            bic <-- Ctr.get b i c;
            whenM (l1 LogicCode.not (LogicCode.equal bic Dom.zero ))
                (seqM (retLoopM (Idx.succ c) (Idx.pred m)
                          (fun k -> Update.update b r c i k) )
                      (Ctr.set b i c Dom.zero)) \} in 
        perform \{
              seqM (retLoopM (Idx.succ r) (Idx.pred n) innerbody) 
                   (Update.update_det brc) \} in
      let dogen a = perform \{
          r <-- Out.R.decl ();
          c <-- retN (liftRef Idx.zero);
          b <-- retN (Ctr.mapper Dom.normalizerf (Ctr.copy a));
          m <-- retN (Ctr.dim1 a);
          n <-- retN (Ctr.dim2 a);
          () <-- Update.D.decl ();
          () <-- Out.P.decl ();
          seqM 
            (retWhileM (LogicCode.and_ (Idx.less (liftGet c) m)
                                       (Idx.less (liftGet r) n) )
               ( perform \{
               rr <-- retN (liftGet r);
               cc <-- retN (liftGet c);
               pivot <-- l1 retN (Pivot.findpivot b rr m cc n);
               seqM (retMatchM pivot (fun pv -> 
                        seqM (zerobelow b rr cc m n pv)
                             (Out.R.succ ()) )
                        (Update.D.zero_sign () ))
                    (Code.update c Idx.succ) \} ))
            (Out.make_result b) \} in
    .<fun a -> .~(runM (dogen .<a>.)) >.
end
\end{code2}
\end{figure}

@@

\subsection*{Acknowledgments}
We wish to thank Cristiano Calgano for his help in adapting camlp4 for
use with MetaOCaml. Many helpful discussions with Walid Taha are very
appreciated.

>>>>

\begin{figure}
\begin{code}
type ('p,'v) monad = 's -> ('s -> 'v -> 'w) -> 'w
    constraint 'p = <state : 's; answer : 'w; ..>

let ret (a :'v) : ('p,'v) monad = fun s k -> k s a
let bind a f = fun s k -> a s (fun s' b -> f b s' k)
let fetch s k = k s s  and  store v _ k = k v ()

let k0 _ v = v
let runM m = fun s0 -> m s0 k0

let l1 f = fun x     -> perform t <-- x; f t
let l2 f = fun x y   -> perform tx <-- x; ty <-- y; f tx ty

@@

let ifL test th el = ret .< if .~test then .~th else .~el >.
let ifM test th el = fun s k -> 
  k s .< if .~test then .~(th s k0) else .~(el s k0) >.
\end{code}
\caption{Our monad}\label{ourmonad}
\end{figure}

@@

In Figure~\ref{ourmonad}, the monad (yielding values of the type |v|)
is implemented as a function of two
arguments: the state (of type |s|) and the continuation. The
continuation receives the current state and the value, and
yields the answer of the type |w|.  The monad is polymorphic over the
three type parameters.  Other implementations are
possible. Except for the code in Figure~\ref{ourmonad}, the rest of our code
treats the monad as a truly abstract data type. The implementation of
the basic
monadic operations |ret| and |bind| is conventional and clearly
satisfies the monadic laws. Other monadic operations
construct computations that do have specific effects.  Operations |fetch| and
|store v| construct computations that read and write the state.

@@

Finally, |runM| runs our monad, that is, given the initial state,
performs the computation of
the monad and returns its result, which in our case is the code
expression. We run the monad by passing it the initial state and the
initial continuation |k0|. We can now re-write our |param_code2|
example of the previous section as |param_code3|.
\begin{code}
let param_code3 plus one =
  let gen x y = bind (retN (plus y one)) (fun ce -> 
                ret (plus ce (plus x ce)))
  in .<fun x y -> .~(runM (gen .<x>. .<y>.) ())>.
\end{code}
% param_code3 plus one;;
%
That does not seem like much of an improvement. With the help of
camlp4 pre-processor, we introduce the |perform|-notation \cite{metamonadsURL},
patterned after the |do|-notation of Haskell. 
\begin{code}
let param_code4 plus one =
  let gen x y = perform ce <-- retN (plus y one);
                        ret (plus ce (plus x ce))
  in .<fun x y -> .~(runM (gen .<x>. .<y>.) ())>.
\end{code}
The function
|param_code4|, written in the |perform|-notation, is equivalent to
|param_code3| -- in fact, the camlp4 preprocessor will convert the
former into the latter. And yet, |param_code4| looks far more
conventional, as if it were indeed in direct style.

\subsection{Generating control statements}
We can write operations that generate code other than let-statements,
e.g., conditionals: see |ifL| in Figure~\ref{ourmonad}. The function |ifL|, 
albeit straightforward, is not as general as we wish: its arguments are
already generated pieces of code rather than monadic values. We
``lift it'', see |ifM'|. 
\begin{code}
let ifM' test th el = perform
  testc <-- test; thc <-- th; elc <-- el;
  ifL testc thc elc
let gen a i = ifM' (ret .<(.~i) >= 0>.) 
                   (retN .<Some (.~a).(.~i)>.) (ret .<None>.)
 in .<fun a i -> .~(runM (gen .<a>. .<i>.) ())>.
\evalresult{.<fun a_1 i_2 -> \protect\newline
let t_3 = (Some a_1.(i_2)) in if (i_2 >= 0) then t_3 else None>.}
let gen a i = ifM (ret .<(.~i) >= 0>.) 
                  (retN .<Some (.~a).(.~i)>.) (ret .<None>.)
 in .<fun a i -> .~(runM (gen .<a>. .<i>.) ())>.
\evalresult{.<fun a_1 i_2 -> 
if (i_2 >= 0) then let t_3 = (Some a_1.(i_2)) in t_3 else None>.}
\end{code}
%
We define functions |l1|,
|l2|, |l3| (analogues of |liftM|, |liftM2|, |liftM3| of Haskell) 
to make such a lifting generic. However we also need
another |ifM| function, with the same
interface (see Figure~\ref{ourmonad}).
The difference between them is
apparent: in the code above with |ifM'|, the let-insertion
happened \emph{before} the if-expression, that is, before the test that
the index |i| is positive. If |i| turned out
negative, |a.(i)| would generate an out-of-bound array access
error. On the other hand, the code with |ifM| accesses the array only
when we have verified that the index is non-negative. This example
makes it clear that the code generation (such as the one in |retN|) is 
truly an effect and we have to be clear about the sequencing of
effects when generating control constructions such as conditionals.
The form |ifM| handles such effects correctly. We
need similar operators for other OCaml control forms: for generating
case-matching statements and |for|- and |while|-loops.

\begin{code}
let loopM low high body = function
  | UP -> (fun s k -> 
      k s .< for j = .~low to .~high do .~(body .<j>. s k0) done >. )
  | DOWN -> (fun s k -> 
      k s .< for j = .~low downto .~high do .~(body .<j>. s k0) done >. )

let whileM cond body = fun s k -> 
  k s .< while .~(cond) do .~(body s k0) done >.

let matchM x som non = fun s k -> k s .< match .~x with
           | Some i -> .~(som .<i>. s k0)
           | None   -> .~(non s k0) >.

let genrecloop gen rtarg = fun s k -> 
  k s .<let rec loop j = .~(gen .<loop>. .<j>. s k0) in loop .~rtarg>.
\end{code}

|loopM| is the analogue of the for-loop.

(* We now define a lot of infrastructure.  We will be quite
   thorough, and define base abstraction and lifted abstractions,
   This will involve a lot of boilerplate code, which unfortunately 
   cannot be so easily automated in MetaOCaml -- that would require 
   introspection.  It could be done in camlp4, but that seems too 
   much as well.  This 'base' could be elided, but when we decide
   to make more use of Abstract Interpretation, we'll regret it,
   so do it now. *)

\begin{code}
let lift x = .< x >.

let liftRef x = .< ref .~x >. 
let liftGet x = .< ! .~x >. 
let unitL = fun s k -> k s .< () >.

let liftPair x = (.< fst .~x >., .< snd .~x >.)

(* logic code combinators - plain and monadic *)
module Logic = struct
  let notL a        = .< not .~a >.
  let equalL a b    = .< .~a = .~ b >.
  let notequalL a b = .< .~a <> .~ b >.
  let andL a b     = .< .~a && .~b >. 
end

(* operations on code indices *)
module Idx = struct
  let zero = .< 0 >.
  let one = .< 1 >.
  let minusone = .< -1 >.
  let succ a = .< .~a + 1 >.
  let pred a = .< .~a - 1 >.
  let less a b = .< .~a < .~b >.
  let uminus a = .< - .~a >.
  let add a b = .< .~a + .~b >.

let cunit = .< () >.
let update a f = let b = f (liftGet a) in .< .~a := .~b >.
let assign a b = .< .~a := .~b >.
let apply  f x = .< .~f .~x >.
let updateM a f = ret (update a f)
let assignM a b = ret (assign a b)
let applyM  f x = ret (apply f x)
\end{code}

maintauining the state

cf Atlas: no unbound variable.
state-changing monad...
extensible object... No worry about init
lots of infrastructure...

mention MLTon.
The reason openrec are so challenging in Ocaml is
because we want to store pieces of code, and exceptions
can't be polymorphic...
The benefit of explicit tags: no need to care about
sharing of modules like Tracking, etc. fewer errors.

In our
case the state is a list (of polymorphic variants), which models an
open discriminated union, as we shall see later.

@@

The simplest parametrization is to make the domain abstract. As it
turns out, we need the following to exist in our domains: $0$, $1$,
$+$, $*$, (unary and binary) $-$, at least \emph{exact} division,
normalization, and potentially a relative size measure. The simplest
case of such domain abstraction is |param_code1|.
There, code-generators such as |plus| and |one|
were passed as arguments. We need far more than
two parameters, so we have to group them. Instead of the grouping
offered by regular records, we use OCaml \emph{structures} (i.e.,
modules)
so we can take advantage of extensibility, type abstraction and constraints,
and especially parameterized structures (\emph{functors}).
We define the type of the domain, the signature |DOMAIN|, which
different domains must satisfy:
\begin{figure}[h]
\begin{code}
type domain_kind = Domain_is_Ring | Domain_is_Field

module S(T:
  sig
	(* Representation type of values,  to be specified *)
    type ('a, 'b) rep
  end) = struct

open T

module type DOMAIN = sig
  type v
  val kind : domain_kind
  val zero : v
  val one : v
  val plus : v -> v -> v
  val times : v -> v -> v
  val minus : v -> v -> v
  val uminus : v -> v
  val div : v -> v -> v
  val better_than : (v -> v -> bool) option
  val normalizer : (v -> v) option
end

module type DOMAINL = sig
  include DOMAIN
  type 'a vc = ('a,v) rep
  val zeroL : 'a vc
  val oneL : 'a vc
  val ( +^ ) : 'a vc -> 'a vc -> 'a vc
  val ( *^ ) : 'a vc -> 'a vc -> 'a vc
  val ( -^ ) : 'a vc -> 'a vc -> 'a vc
  val uminusL : 'a vc -> 'a vc
  val divL : 'a vc -> 'a vc -> 'a vc
  val better_thanL : ('a vc -> 'a vc -> ('a,bool) rep) option
  val normalizerL : ('a vc -> 'a vc) option
end

module IntegerDomain = struct
    type v = int
    let kind = Domain_is_Ring
    let zero = 0
    let one = 1
    let plus x y = x + y
    let times x y = x * y
    let minus x y = x - y
    let uminus x = -x
    let div x y = x / y
    let normalizer = None
    let better_than = Some (fun x y -> abs x > abs y)
end

module T = Domains_sig.S(struct type ('a, 'b) rep = ('a, 'b) code end)
open T

open Domains_common
(* because the operations are "syntactic" to a certain extent,
   we have to repeat ourselves a lot *)
module IntegerDomainL = struct
    include IntegerDomain
    type 'a vc = ('a,v) code
    let zeroL = .< 0 >.  
    let oneL = .< 1 >. 
    let (+^) x y = .<.~x + .~y>. 
    let ( *^ ) x y = .<.~x * .~y>.
    let ( -^ ) x y = .<.~x - .~y>.
    let uminusL x = .<- .~x>.
    let divL x y = .<.~x / .~y>. 
    let normalizerL = None
    let better_thanL = Some (fun x y -> .<abs .~x > abs .~y >. )
end

\end{code}
\end{figure}

@@

\begin{code2}
(* The `keyword' list of all the present internal features *)
module type INTERNAL_FEATURES = sig
  module R      : TrackRank.RANK
  module P      : TRACKPIVOT
  module L      : LOWER
end

module type OUTPUT = functor(OD : OUTPUTDEP) -> sig
  module IF : INTERNAL_FEATURES
  type res
  val make_result : 'a wmatrix ->
   (<classif : 'a; 
     state : [> `TDet of 'a OD.Det.lstate |
                'a IF.R.tag_lstate |
                `TPivot of 'a IF.P.lstate |
                `TLower of 'a IF.L.lstate]; ..>,res) cmonad
   (*
    ('a,res,[> 'a OD.Det.tag_lstate | 'a IF.R.tag_lstate 
             | 'a IF.P.tag_lstate   | 'a IF.L.tag_lstate],'w) cmonad
      *)
end
module OutDetRank(OD : OUTPUTDEP) = struct
  module IF = struct
      module R   = Rank
      module P   = DiscardPivot
      module L   = NoLower end
  type res = C.contr * C.Dom.v * int
  let make_result m = perform
    det  <-- OD.Det.fin ();
    rank <-- IF.R.fin ();
    ret (Tuple.tup3 m.matrix det rank)
  (* Initialization: check the preconditions of instantiation of this struct*)
  let _ = OD.Det.fin ()
  let _ = IF.R.fin ()
end
\end{code2}

@@

\begin{code}
(* The `keyword' list of all the present external features *)
module type FEATURES = sig
  module Det       : DETERMINANT
  module PivotF    : PIVOT
  module PivotRep  : PIVOTKIND
  module Update    : UPDATE
  module Input     : INPUT
  module Output    : OUTPUT
end
\end{code}

\begin{figure}
\begin{code2}
module GenGE(F : FEATURES) = struct
    module O = F.Output(F)

    let wants_pack = O.IF.L.wants_pack
    let can_pack   = 
        let module U = F.Update(F.Det) in
        (U.upd_kind = DivisionBased)
    (* some more pre-flight tests *)
    let _ = ensure ((not wants_pack) || can_pack) 
           "Cannot return a packed L in this case"

    let zerobelow mat pos = 
        let module IF = O.IF in
        let module U = F.Update(F.Det) in
        let innerbody j bjc = perform
            whenM (Logic.notequalL bjc C.Dom.zeroL ) (perform
                det <-- F.Det.get ();
                optSeqM (Iters.col_iter mat.matrix j (Idx.succ pos.p.colpos) 
               (Idx.pred mat.numcol) C.getL
                      (fun k bjk -> perform
                      brk <-- ret (C.getL mat.matrix pos.p.rowpos k);
                      U.update bjc pos.curval brk bjk 
                          (fun ov -> C.col_head_set mat.matrix j k ov) det) UP )
                      (IF.L.updt mat.matrix j pos.p.colpos C.Dom.zeroL 
                          (* this makes no sense outside a field! *)
                          (C.Dom.divL bjc pos.curval))) in
        perform
              seqM (Iters.row_iter mat.matrix pos.p.colpos
              (Idx.succ pos.p.rowpos)
              (Idx.pred mat.numrow) C.getL innerbody UP) 
                   (U.update_det pos.curval)

   let init input = perform
        let module IF = O.IF in
          (a,rmar,augmented) <-- F.Input.get_input input;
          r <-- IF.R.decl ();
          c <-- retN (liftRef Idx.zero);
          b <-- retN (C.mapper C.Dom.normalizerL (C.copy a));
          m <-- retN (C.dim1 a);
          rmar <-- retN rmar;
          n <-- if augmented then retN (C.dim2 a) else ret rmar;
          F.Det.decl ();
          IF.P.decl rmar;
          _ <-- IF.L.decl (if wants_pack then b else C.identity rmar m);
          let mat = {matrix=b; numrow=n; numcol=m} in
          ret (mat, r, c, rmar)

   let forward_elim (mat, r, c, rmar) = perform
        let module IF = O.IF in
          whileM (Logic.andL (Idx.less (liftGet c) mat.numcol)
                              (Idx.less (liftGet r) rmar) )
             ( perform
             rr <-- retN (liftGet r);
             cc <-- retN (liftGet c);
             let cp  = {rowpos=rr; colpos=cc} in
             let module Pivot = F.PivotF(F.Det)(IF.P) in
             pivot <-- l1 retN (Pivot.findpivot mat cp);
             seqM (matchM pivot (fun pv -> 
                      seqM (zerobelow mat {p=cp; curval=pv} )
                           (IF.R.succ ()) )
                      (F.Det.zero_sign () ))
                  (updateM c Idx.succ) )

   let gen input = perform
          (mat, r, c, rmar) <-- init input;
          seqM 
            (forward_elim (mat, r, c, rmar))
            (O.make_result mat)
end
\end{code2}
\end{figure}

@@

\subsection*{Acknowledgments}
We wish to thank Cristiano Calgano for his help in adapting camlp4 for
use with MetaOCaml. Many helpful discussions with Walid Taha are very
appreciated. The implementation of the monadic notation, |perform|,
was the joint work with Lydia van Dijk.
